---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r Preprocessing, echo=TRUE}
#Create new working directory for this project
wd <- getwd() 
if (!file.exists("./Project1")) {dir.create("./Project1")}
newWD <- paste(wd,"/Project1", sep="")
setwd(newWD)

#get source data file from my local repo clone
fileURL <- "~/DataScience/datasciencecoursera/RepData_PeerAssessment1/activity.zip"

#unzip file and save to new working dir
unzip(fileURL, exdir = newWD, overwrite=TRUE)

#read data
activityDataRaw <- read.csv("activity.csv")
dim(activityDataRaw) # This is to confirm we have the expected 
head(activityDataRaw)
```
This output confirms that we have the expected data as indicated
in the instructions.  

## What is mean total number of steps taken per day?
```{r MeanSteps, echo=TRUE}
# 1. Calculate the total number of steps taken per day
# I decided simply to use base graphics for this task. Nothing fancy.
dailySteps <- tapply(activityDataRaw$steps, activityDataRaw$date, sum,
                     na.rm = TRUE) #we were told to ignore NAs for this part

# 2. Make a histogram of the total number of steps taken each day
hist(dailySteps, main="Histogram of Daily Step Totals", 
     xlab="Total Daily Steps", col="lightblue", ylim=c(0, 30))

# 3. Calculate and report the mean and median 
# of the total number of steps taken per day
meanDailySteps <- mean(dailySteps)
medianDailySteps <- median(dailySteps)

mmPlot <-
    barplot(c(meanDailySteps, medianDailySteps), 
            main="Daily Steps: Comparison of Central Measures",
            ylim=c(0, 12000), names.arg=c("Mean Daily Steps", "Median Daily Steps"),
            ylab="Number of Daily Steps", col=c("lightblue", "pink"))
# Now add the y values as text
text(y=c(meanDailySteps, medianDailySteps), 
     x=mmPlot, 
     pos=3, #Place data labels above the bars
     labels=as.character(c(round(meanDailySteps, digits=1), medianDailySteps)))

```  

Notice in plot that the median is almost 1041 steps more than the mean.

## What is the average daily activity pattern?
```{r DailyPatternInstructions, echo=TRUE}
# 1. Make a time series plot (i.e. type = "l" ) of the 5-minute interval (x-axis) 
#    and the average number of steps taken, averaged across all days (y-axis)
# 2. Which 5-minute interval, on average across all the days in the dataset, 
#    contains the maximum number of steps?
```

Question 1 is slightly confusing, but here is my interpretation of it:
He is asking us to compute the average number of steps taken for each 5-min
interval.  Since there are 288 5-minute samples in a 
day, we will have 288 data points, each one being the mean for that same
interval across all 61 days (fewer days if some are NA).
```{r DailyPatternCode, echo=TRUE}
# Question 1 Make a time series plot of the 5-minute interval (x-axis) 
# and the average number of steps taken, averaged across all days (y-axis)
intervalMeanSteps <- tapply(activityDataRaw$steps, activityDataRaw$interval, 
                     mean, na.rm = TRUE)
plot(intervalMeanSteps, type="l", 
     main="Mean Steps of 288 5-minute Samples\n(missing data omitted)",
     xlab="5-minute Interval", ylab="Mean Steps across 61 Days", col="blue",
     xlim=c(0, 300))

# Question 2 Which 5-minute interval, on average across all the days 
# in the dataset, contains the maximum number of steps?
```  

intervalMeanSteps has length 288 and each value is the mean of the
steps for that interval across all days, as we can see here:
```{r , echo=TRUE}
length(intervalMeanSteps)
```
```{r}
head(intervalMeanSteps)
```

So now we just have to find which of those intervals has the highest value.
The highest value is:
```{r}
max(intervalMeanSteps)
```
So let's find out the index of that interval
```{r}
indexOfmaxMeanStepInterval <- which(intervalMeanSteps==max(intervalMeanSteps))
indexOfmaxMeanStepInterval
```

So the 104th element of intervalMeanSteps refers to interval 835
Looking at the plot, this appears correct since the highest point looks
slightly above 200 on the y-axis and slightly higher than 100 on the x-axis.

That is our answer, but to double-check its correctness:

```{r}
intervalMeanSteps[indexOfmaxMeanStepInterval[[1]]]
```

I.e. this is correct.
So the answer to question 2 is that interval 835, on average, has the 
highest number of steps: 206.1698

## Imputing missing values
```{r ImputingNA, echo=TRUE}
# 1. Calculate and report the total number of missing values in the dataset 
#    (i.e. the total number of rows with NA s)

missingSteps <- is.na(activityDataRaw$steps)
sum(missingSteps) #This shows our answer.  
```

We can verify this independently by calling the following,
which shows the number of NAs at the lower left of the output:
```{r ConfirmNAs, echo=TRUE}
summary(activityDataRaw)

# 2.  Devise a strategy for filling in all of the missing values in the dataset. 
# 3.  Create a new dataset that is equal to the original dataset but with the 
#     missing data filled in.
```
I decided to round the mean for each 5-min interval to the nearest whole number
e.g. for interval 1840, the mean is 85.33962, so the round() of that is 85,
which comes from round(intervalMeanSteps[["1840"]])

```{r FillInNAs, echo=TRUE}
# get row indices of raw data with missing steps
indicesOfNA <- which(missingSteps, arr.ind=TRUE)

# Define function on data frame and indices into that df
# to replace the NA vals in the 'steps' var with the round of the steps
# in the same interval using 'intervalMeanSteps', previously computed.
fillInNA <- function(df, indices) {
    # replace each NA in 'steps' var with the nearest whole number
    # from the corresponding interval previously computed in intervalMeanSteps
    for (i in indices) {
        intervalOfNA <- df[i, "interval"]
        #next line references intervalMeanSteps as a global
        df$steps[i] <- round(intervalMeanSteps[[as.character(intervalOfNA)]])
    }
    df
}

activityDataFilled <- fillInNA(activityDataRaw, indicesOfNA)

# 4. Make a histogram of the total number of steps taken each day and Calculate 
#    and report the mean and median total number of steps taken per day. 
#    Do these values differ from the estimates from the first part of the assignment? 
#    What is the impact of imputing missing data on the estimates 
#    of the total daily number of steps?

dailyStepsFilled <- tapply(activityDataFilled$steps, activityDataFilled$date, 
                           sum) 

#Make a histogram of the total number of steps taken each day
hist(dailyStepsFilled, main="Histogram of Daily Step Totals (filled NAs)", 
     xlab="Total Daily Steps (filled NAs)", col="blue", ylim=c(0, 40))

#Instructions: Calculate and report the mean and median 
#of the total number of steps taken per day
meanDailyStepsFilled <- mean(dailyStepsFilled)
medianDailyStepsFilled <- median(dailyStepsFilled)

mmPlotFilled <-
    barplot(c(meanDailyStepsFilled, medianDailyStepsFilled), 
            main="Daily Steps (filled NAs): Comparison of Central Measures",
            ylim=c(0, 12000), 
            names.arg=c("Mean Daily Steps", "Median Daily Steps"),
            ylab="Number of Daily Steps", col=c("blue", "red"))
# Now add the y values as text
text(y=c(meanDailyStepsFilled, medianDailyStepsFilled), 
     x=mmPlotFilled, 
     pos=3, #Place data labels above the bars
     labels=as.character(c(round(meanDailyStepsFilled, digits=1), 
                           medianDailySteps)))

# Now notice how, with filled NA data, the mean and median are nearly identical

#Obviously, imputing the missing values indeed makes a huge difference
#in the graph compared to the raw numbers.  With imputed values, 
#the mean and median are nearly identical, whereas with the raw data
#the difference was almost 1041 steps.  
#And also the shape of the 2 histograms is VERY different in raw vs. imputed,
#as is the absolute height of the bars (which would be expected).
```


## Are there differences in activity patterns between weekdays and weekends?
```{r ActivityPatterns, echo=TRUE}
# For this part the weekdays() function may be of some help here. Use the dataset 
# with the filled-in missing values for this part.
# 1. Create a new factor variable in the dataset with two levels – “weekday” and 
#    “weekend” indicating whether a given date is a weekday or weekend day.
# 2. Make a panel plot containing a time series plot (i.e. type = "l" ) 
#    of the 5-minute interval (x-axis) and the average number of steps taken, 
#    averaged across all weekday days or weekend days (y-axis). See the README file 
#    in the GitHub repository to see an example of what this plot should look 
#    like using simulated data.

isWeekend <- function(dateAsCharacter) {
                weekdays(as.Date(dateAsCharacter)) %in% c("Saturday", "Sunday")
                }
is.weekend <- isWeekend(activityDataFilled$date)

is.weekend[is.weekend==TRUE] <- "weekend"
is.weekend[is.weekend==FALSE] <- "weekday"
        
activityDataFilled$TypeOfDay <- as.factor(is.weekend)

# Now make the panel plot and I'll use ggplot2

library(ggplot2)

tsPlot <- ggplot(activityDataFilled, aes(interval, steps))
tsPlot + facet_grid(TypeOfDay ~ .) +
    xlab("Interval") +
    ylab("Number of steps") +
    ggtitle("Time Series Comparison of Mean Steps: Weekday vs. Weekend") +
    stat_summary(fun.y="mean", geom=c("line"), color="blue")
```
